\chapter{Introduction}

Developing professional software these days has become quite an involved affair.
Not long ago, a team of engineers would sit down to develop an application that
would simply and modestly run on a single computer. Such software would operate
completely in its own world, blissfully unaware of the internet, only making a
network call on seldom occasions, \eg to phone home to its vendor to ask for
software updates. This was the state of software development a few short years
ago.

% have woven large swaths of themselves

Today, large swaths of most applications have been woven into ``the cloud'' or
other network services. Web applications are becoming patchwork quilts of made
up of calls to multitudes of different microservices. Modest mobile apps now
make network calls to dozens or even hundreds of services. Meanwhile as software
becomes evermore pervasive, weaving itself more into more of our daily habits in
more places, content providers are focusing their energies on collecting any and
all seemingly innocuous pieces of our data that they can, in an attempt to
unlock some sort of market value in peoples' trails of digital breadcrumbs. With
all of this data piling up, industry and academia are scrambling to build
distributed systems that can help more users make sense of it--clusters of
machines working together to churn through datasets too large to fit in the
memory of a single machine.

This is the new computing landscape; the network has become ubiquitous and is
now baked into much of the programming that professional developers do.

Meanwhile, at the same time, we're witnessing a renaissance of functional
programming so prevalent that it has permeated the daily routines of software
developers on all ends of the the software development spectrum, from the
client-side\footnote{Popular functional languages for the client-side include:
numerous JavaScript libraries such as
\href{http://underscorejs.org}{Underscore.js}, Elm~\cite{Elm},
\href{http://www.purescript.org/}{PureScript},
\href{http://www.scala-js.org/}{Scala.js}, amongst many others.} to the
server-side.\footnote{Popular functional languages for the server-side include:
Scala~\cite{Odersky10}, Clojure~\cite{Clojure}, Erlang~\cite{Erlang},
Haskell~\cite{Haskell}, amongst many others.} Further, the distributed system
core of services like Twitter are based on functional APIs~\cite{Eriksen2013},
and frameworks for big data analytics like Spark~\cite{Spark} credit functional
patterns for enabling more powerful computation patterns; \ie general graphs of
computations built up of compositions of higher order combinators rather just
maps and reduces like in MapReduce~\cite{MateiScalaByTheBay}. Just about
everywhere you look now a days, you'll find functionally-inspired software
springing up in the wild.


% , and that
% style of development is becoming increasingly functional.

But how have our most important tools in professional software development --
programming languages -- kept up as the network and functional programming have
begun to proliferate software development environments?

As it turns out, there are still numerous issues using language constructs such
as objects and functions in a distributed setting. Moreover, due to their nature
of being built-in to the language, it is impossible to rely on libraries and
frameworks to provide support for the reliable distribution of these constructs.
As a result, even mature libraries and frameworks can exhibit bugs that are hard
to diagnose and fix.

For example, in mainstream languages like Java, even the serialization of simple
objects, a prerequisite for sending them across the network, can lead to runtime
errors that can be difficult to diagnose and fix. While alternative
serialization frameworks exist, they feature only rudimentary language
integration, in stark contrast to native platform support.

This issue is exacerbated when using closures, which are increasingly appearing
in popular distributed frameworks such as Spark~\cite{Spark} and
Scalding~\cite{Scalding}. One of the main reasons is that closures, as they
exist in virtually all wide-spread languages, leave essential components, such
as their captured variables, implicit, preventing customizations necessary to
make closures safer and more efficient to distribute.

\begin{itemize}
  \item {\bf \em Pickling (serialization)}
  \item {\bf \em Functions}
  \item {\bf \em Asynchronous programming}
\end{itemize}

% Researchers and practitioners have explored different approaches to make dist.
% prog. simpler, more reliable, and more performant. (1) message passing, (2) ..., ...
%
% The majority of widely-used PLs relegate distribution to libraries. While many
% of the above-mentioned approaches have been realized as libraries for
% widely-used languages, there are still numerous issues using language constructs
% such as objects and functions in a distributed setting. Moreover, due to their
% nature of being built-in to the language, it is impossible to rely on libraries
% and frameworks to provide support for the reliable distribution of these
% constructs. As a result, even mature libraries and frameworks can exhibit bugs
% that are hard to diagnose and fix (or suffer from the limitations of their
% underlying languages/runtimes).
%
% For example, in mainstream languages like Java, even the act of serializing
% simple objects, a prerequisite for sending them across the network, can lead to
% runtime errors that can be difficult to diagnose and fix. While alternative
% serialization frameworks exist, they feature only rudimentary language
% integration, in stark contrast to native platform support.
%
% This issue is exacerbated when using closures, which are increasingly appearing
% in popular distributed frameworks such as Spark~\cite{Spark} and
% Scalding~\cite{Scalding}. One of the main reasons is that closures, as they
% exist in virtually all wide-spread languages, leave essential components, such
% as their captured variables, implicit, preventing custom treatment necessary to
% make closures safer and more efficient to distribute.
%
% [connect]
%
% Scala is a functional programming language that runs on the
% JVM\footnote{Although Scala now also targets JavaScript with a Scala-to-JS
% compiler known as \href{http://www.scala-js.org/}{Scala.js}} which features both a rich object model as well as extensive support for functional programming.
% Scala delegates aspects such as serialization to its underlying platform, typically Java. As a result, Scala developers are transitively also affected by issues of Java's built-in serialization. Functions too bring a whole host of complications wh
%
% Given that we desire solutions to issues traditionally impossible to fix using libraries alone, the question arises how to best extend languages in a way that does not require intrusive changes to their specification, compilers, and tools.
%
% A goal of this dissertation is to revisit the fundamental concepts of modern
% languages, objects and functions, and to make them safer and more efficient to
% use in a distributed environment. We focus on three important and orthogonal
% building blocks for distributed programming:

% This is
% where Scala shines.

% Languages themselves have actually done little
%
% Little has actually been done in recent years in the programming languages
% community to address this fundamental shift.
% % Instead, the focus of the effort
% % seems to be on.
%
% Scala is a functional programming language that runs on the
% JVM\footnote{Although Scala now targets JavaScript too with a Scala-to-JS
% compiler known as \href{http://www.scala-js.org/}{Scala.js}}. Scala is popular
% amongst developers of distributed systems for two prime reasons; (1) as a JVM
% language, it seamlessly interoperates with Java, the de facto language of choice
% amongst developers of big data systems, and (2) its functional features like
% closures combined with a concise
%
% Functional language features make .
%
% % distributed programming is now the norm, it's in every program these days.
%
% but how have languages kept up? well, they really haven't. instead there are a
% handful of disparate attempts at making distributed programming easier from a
% higher level (handling concerns like consistency etc)
%
% Meanwhile, functional programming is also on the rise, gaining popularity as
% functional languages begin to mainstream.
%
%
% furthermore, we propose a new dist. prog. model that at the same time (a)
% embraces these improved building blocks and (b) generalizes existing, well-known
% prog.models.
%
% the ultimate goal is to simplify the lives of both users and developers of dist.
% systems
%
% we believe the ideas and systems in this diss. are steps towards this goal; this
% is also witnessed by strong interest from industry in some of these systems
%
%
%  Recently, a
%
%  rash of distributed systems have
% sprung up, each promising to unlock information hidden in these mountains of
% messy data. Here too,
%
% The internet and other the network is
% now baked into most of the programming that we do.
%
%
% Distributed programming is fundamentally different than . Without even realizing
% it,. It requires a
% To date, no mainstream programming language has been designed with distribution in mind since the start. ...No...
%
% Programming for distributed systems is hard. All it takes is a quick glance at
% the literature, both academic and industrial, to see that distributed systems
% builders have long since grappled with fundamental concerns like consistency,
% availability, and latency across systems designs.
%
% The performance characteristics of the underlying programming language and
% runtime, however, should not weigh in
%
% Major goal of this thesis is to retroactively add support for distributed
% computation to an existing and widespread language with functional features.
%
% Many efforts have been undertaken to design from scratch a new foundation for
% distribution. Cite Leo.
%
%
% The closest the literature gets at present to the intersection of programming
% languages and falls in the direction of Datalog and the idea that ensuring that
% properties like monotonicity
%
% Nitty gritty concerns like serialization or ensuring that are left largely
% unconsidered. As an industry, we simply deal with the fact that Java 8 now has
% closures, but that closures have no added support to ensure that they are
% serializable, instead suggesting ``the serialization of lambda expressions is
% strongly discouraged.''\footnote{From the official Java documentation on lambda
% expressions, available at
% \url{https://docs.oracle.com/javase/tutorial/java/javaOO/lambdaexpressions.html\#serialization}}
%
% \section{Distributing Computation}
%
%
%
% Erlang, Elixir. But no types. Go for fault tolerance? Bloom language.
%
% Sequoia. (?)
%
% Oz dataflow.
%
% Up to 50\% of the time for some Spark jobs spent in
% serialization~\cite{OusterhoutSerialization}.
%
% Datalog, Datomic, Cascalog. Query evaluation with Datalog is based on first
% order logic, and is thus sound and complete. However, Datalog is not Turing
% complete, and is thus used as a domain-specific language that can take advantage
% of efficient algorithms developed for query resolution. Datalog is a lightweight
% deductive database system where queries and database updates are expressed in
% the logic language. The use of Datalog syntax and an implementation based on
% tabling intermediate results ensures that all queries terminate.
%
% Datalog is a nonprocedural query language based on the logic-programming
% language Prolog.  A user describes the information desired without giving a
% specific procedure for obtaining that information. Datalog simplifies writing
%
% Much effort invested in a formal foundation for concurrent and distributed
% programming -- e.g., process calculi. Pi calculus. Join calculus.
%
% Join-patterns provides a way to write concurrent, parallel and distributed
% computer programs by message passing. Compared to the use of threads and locks,
% this is a high level programming model using communication constructs model to
% abstract the complexity of concurrent environment and to allow scalability. Its
% focus is on the execution of a chord between messages atomically consumed from a
% group of channels.
%
% Recent research has explored using Datalog-based languages to express a
% distributed system as a set of logical invariants.
%
% Languages foundations like Dedalus~\cite{Dedalus} focus on providing a minimal set of
% primitives for programming and reasoning about distributed systems. Dedalus
% reduces to a subset of Datalog with negation, aggregate functions,
% successor and choice, and admits an explicit representation of time into the
% logic language in order to provide a declarative foundation for the two
% signature features of distributed systems: mutable state, and asynchronous
% processing and communication.
%
% NoSQL = Datalog! Function-passing is a stateless model of this idea. All about
% using logic to reduce coordination.
%
% These languages and efforts differ from the effort presented in this thesis in
% that logic-based . They are trying to figure out ways to jointly do
% optimizations and to ensure a number of other properties important to
% distributed systems with state remain in tact. When a theoretical result is
% arrived at in this work of databases and data-centric programming systems, the
% sole effort to date concerned with integrating these advancements into the
% programming language has focused on providing a restricted logic programming
% language to attempt to ensure correctness and ease of reasoning. These are not
% the goals of this thesis. Instead, we focus less on providing langauge featuers designed to shield users from complications that arise when having to worry about consistency (after all,
%
%
% The rationale of some in purporting a language to be more adept at distributed
% programming has focused on languages which remove limitations and which enable
% more features to be distributed.
%
% That is not the central tenet of this thesis. Rather, this thesis argues that
% while indeed more language features ought to be more reliably distributed (e.g.,
% function closures) it might not be preferable that all features be given
% unrestricted support for distribution. That is, in this thesis,
%
% This thesis follows the advice given in 's Note On Distributed computing, where
% important advice is to avoid abstracting over network communication. Other
% models tout such abstraction to be a feature, or to from a programming language
% perspective provide for a more powerful or otherwise desirable . Like Note, we
% argue that this is not the case. Designers of distributed systems must have a
% handle on and be aware of all situations that could cause network communication.
%
% The distributed model of computation behind Oz~\cite{DistributedOz} abstracts
% over communication boundaries. Network transparency
% ~\cite{ConceptsTechniquesModelsProgramming}
%
% In this thesis, rather than abstracting over network communication, we provide
% firmer, more desirable primitives to initiate your own network communication.


\section{Thesis}

The goal of this dissertation is to improve the above state of affairs
Part of this thesis aims to imagine what a more functional model for distributed programming looks like.

This dissertation seeks to improve the above state of affairs from the
perspective of

This dissertation addresses the following questions:

\begin{itemize}

	\item How can existing programming language features be improved in order to
	better support concerns like performance and latency across a general slice of
	distributed systems?

	\item Which important features and aspects of existing programming languages
	are left unsupported by the language in the face of distribution? Is it
	possible to support such features?

	\item How can core ideas behind the development of functional programming be
	applied to the distributed scenario? What other models for functional
	programming in a distributed environment are there?

	\item What are good abstractions for reasoning about concerns like network I/O
	and failure at the level of the compiler and programming language?

\end{itemize}

\section{Contributions}

This dissertation describes a number of extensions and libraries in and for
Scala which aim to provide a more reliable foundation for building distributed
systems atop of. Importantly, a goal throughout this dissertation was not only
to add better support the basics of distributed programming, but to do it {\em
without extending the Scala compiler}\footnote{Finding ways mainstream language with
hundreds of thousands of users.}.

\begin{itemize}

	\item Abstractions and data structures for asynchronous programming.

	\item

\end{itemize}

\section{Structure}

The rest of this dissertation is organized as follows.

\begin{itemize}

	\item Chapter 2 describes {\em futures} and {\em FlowPools}, funtionally-inspired
	and fully asynchronous and non-blocking single-assignment variables (futures)
	and pools (FlowPools) useful for reducing coordination in distributed systems.
	The chapter sketches a proof of linearizability, lock-freedom, and determinism
	of FlowPools. The full proof of lock-freedom can be found in
	Appendix~\ref{appdx:flowpools}, and the full proofs of linearizability and
	determinism can be found in the companion technical report~\cite{Prokopec12}.

	\item Chapter 3 introduces {\em object oriented picklers} and {\em
	scala/pickling}, a new distribution-focused approach to serialization that
	generates serialization code statically, allowing for more type safety. The
	chapter includes a formalization of object oriented picklers as well as a
	description of the generation algorithm used for automatically generating
	picklers for arbitrary types. A performance evaluation is also included which
	examines the performance of the serialization framework across different sorts
	of serilaization workloads, and which compares scala/pickling against other
	state-of-the art serialization systems like Java and Kryo, and reports
	significant speedups.

	\item Chapter 4 covers a new technique for extensible and static datatype
	generic programming. In this chapter, the generation technique used for
	generating pickling code is generalized to be able to generate arbitrary type
	class instances, at compile time.

	\item Chapter 5 introduces {\em spores}, a new abstraction and type system
	designed to enable function closures to be serializable by design. The type
	system presented here also generalizes its added static checking capabilities
	to arbitrary user-defined {\em properties}, \eg immutability.

	\item Chapter 6 describes a new programming model for functional distributed
	programming called {\em function-passing} which aims to simplify the
	implementation and reasoning of fault tolerance mechanisms. Because the model
	requires user programs to themselves be stateless, issues like consistency are
	moot. This programming model can be thought of as a generalization of the Spark
	or MapReduce programming model.

	\item Chapter 7 concludes and discusses possible directions for future work.

\end{itemize}

\section{Previously Published Material}

This dissertation draws heavily on earlier work described in the following
papers, written jointly with several collaborators (in the order of appearance
in this disseratation):

\begin{itemize}

	\item Prokopec, Miller, Schlatter, Haller, and Odersky (2012).
	\href{http://infoscience.epfl.ch/record/180265}{FlowPools: A lock-free
	deterministic concurrent dataflow abstraction}. In proceedings of Languages and
	Compilers for Parallel Computing (LCPC).

	\item Miller, Haller, Burmako, and Odersky (2013).
	\href{http://infoscience.epfl.ch/record/188383}{Instant Pickles: Generating
	object-oriented pickler combinators for fast and extensible serialization}. In proceedings of the ACM SIGPLAN International Conference on Object-Oriented Programming, Systems, Languages, and Applications (OOPSLA).

	\item Miller, Haller, and Odersky (2014).
	\href{http://infoscience.epfl.ch/record/191239}{Spores: A type-based
	foundation for closures in the age of concurrency and distribution}. In
	proceedings of the European Conference on Object-Oriented Programming (ECOOP).

	\item Haller and Miller (2015).
	\href{https://infoscience.epfl.ch/record/205039}{Distributed Programming
	via Safe Closure Passing}. In proceedings of Programming Language Approaches to
	Concurrency and Communication-centric Software (PLACES).

\end{itemize}

Works that this dissertation draws upon that have been submitted but at the time
of the writing remain in technical report form include:

\begin{itemize}

	\item Miller, Haller, and C. D. S. Oliveira (2015). \href{http://infoscience.epfl.ch/record/199389}{Self-Assembly: Lightweight language extension and datatype generic programming, all-in-one!}. EPFL technical report \#EPFL-CONF-199389.

	\item Miller and Haller (2015).
	\href{http://infoscience.epfl.ch/record/205822}{Function Passing: A model for
	typed, distributed functional programming}. EPFL technical report
	\#EPFL-CONF-205822.

\end{itemize}
